---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ssdsims

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/poissonconsulting/ssdsims/actions/workflows/R-CMD-check.yml/badge.svg)](https://github.com/poissonconsulting/ssdsims/actions/workflows/R-CMD-check.yml)
<!-- badges: end -->

ssdsims runs reproducible simulation studies for species sensitivity
distribution (SSD) models built on the
[ssdtools](https://bcgov.github.io/ssdtools/) package. You describe a study
**declaratively** as a *scenario*; ssdsims expands it into per-step task tables,
draws the data, fits the distributions, and estimates the hazard
concentrations. The same scenario runs three ways over one execution core ---
an in-memory baseline, single-core Hive-partitioned Parquet *shards*, and a
[targets](https://docs.ropensci.org/targets/)-based shard pipeline for running
in parallel or on a cluster --- with byte-identical, reproducible per-task
results throughout.

New here? Start with
[`vignette("ssdsims")`](https://poissonconsulting.github.io/ssdsims/articles/ssdsims.html)
for the overview and a recommended reading order.

## Installation

Install the development version from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("poissonconsulting/ssdsims")

# or
# install.packages("remotes")
remotes::install_github("poissonconsulting/ssdsims")
```

## Usage

Assemble the data, declare a scenario, and run the in-memory baseline:

```{r}
library(ssdsims)

scenario <- ssd_define_scenario(
  ssd_scenario_data(ssddata::ccme_boron),
  nsim = 2L,
  seed = 42L,
  nrow = c(6L, 10L),
  ci = TRUE,
  nboot = 10L,
  ci_method = c("multi_fixed", "weighted_samples")
)
scenario

ssd_run_scenario_baseline(scenario)
```

The scenario is purely declarative --- it draws no random numbers and depends on
no `targets` --- so the same object drives every runner. See
[`vignette("defining-a-scenario")`](https://poissonconsulting.github.io/ssdsims/articles/defining-a-scenario.html)
for the scenario object and
[`vignette("sharded-pipeline")`](https://poissonconsulting.github.io/ssdsims/articles/sharded-pipeline.html)
for the shard and `targets` pipeline.

## Capability map

| Area | Article |
|---|---|
| Define a scenario and expand it into `sample`/`fit`/`hc` task tables | [`vignette("defining-a-scenario")`](https://poissonconsulting.github.io/ssdsims/articles/defining-a-scenario.html) |
| Run as Hive-partitioned Parquet shards and a `targets` pipeline | [`vignette("sharded-pipeline")`](https://poissonconsulting.github.io/ssdsims/articles/sharded-pipeline.html) |
| Combine several scenarios into one ragged design | [`vignette("scenario-to-design")`](https://poissonconsulting.github.io/ssdsims/articles/scenario-to-design.html) |
| Run on a SLURM (or other) cluster | [`vignette("cluster-pipeline")`](https://poissonconsulting.github.io/ssdsims/articles/cluster-pipeline.html) |
| Ship shards to cloud object storage | [`vignette("cloud-upload")`](https://poissonconsulting.github.io/ssdsims/articles/cloud-upload.html) |
| Estimate compute cost before a run | [`vignette("cost-estimation")`](https://poissonconsulting.github.io/ssdsims/articles/cost-estimation.html) |
| Analyse observed compute cost after a run | [`vignette("cost-analysis")`](https://poissonconsulting.github.io/ssdsims/articles/cost-analysis.html) |

Reproducible per-task RNG (the dqrng `pcg64` backend with per-task primers)
underpins all of the above; see
[`task_primer()`](https://poissonconsulting.github.io/ssdsims/reference/task_primer.html).
