---
title: "Output correctness — `ssdsims` targets experiments"
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(duckplyr)
  library(readr)
  library(purrr)
})

summary_csv <- "../results/summary.csv"
stopifnot(file.exists(summary_csv))
sm <- readr::read_csv(summary_csv, show_col_types = FALSE)
```

Twelve experiment runs wrote bootstrap HC results to
`data/<config_id>/branch-*.parquet`. Each config should produce the same
*combined* answer regardless of how many branches it was split across —
this notebook checks that.

## Per-config row counts

```{r}
counts <- sm |>
  rowwise() |>
  mutate(
    files = list(list.files(file.path("..", "data", config_id),
                            pattern = "\\.parquet$", full.names = TRUE)),
    n_rows = {
      f <- files
      if (length(f) == 0) NA_integer_ else
        duckplyr::read_parquet_duckdb(f) |>
          summarise(n = n()) |> collect() |> pull(n)
    }
  ) |>
  ungroup() |>
  select(size, split, n_branches, n_atomic_units, n_rows)

counts |> knitr::kable()
```

Each atomic unit is one `ssd_hc()` call. With `est_method = "multi"` and
`samples = TRUE`, every `(nrow, ci_method, nboot, proportion, sim)` pair
contributes one row to the output, so we expect

```
n_rows == n_atomic_units * nsim
```

with `nsim = 2` for every config:

```{r}
counts |>
  mutate(
    expected = n_atomic_units * 2L,
    delta    = n_rows - expected
  ) |>
  select(size, split, n_rows, expected, delta) |>
  knitr::kable()
```

If `delta` is zero everywhere, splitting is row-count-preserving.

## Cross-split agreement

For each `size`, the `est`/`lcl`/`ucl` columns at every
`(sim, nrow, ci_method, nboot, proportion)` key should match across the
four splits, because seeds are deterministic and the work per atomic unit
is identical regardless of how it's batched.

```{r}
read_all <- function(size_val) {
  rows <- sm |> filter(size == size_val)
  purrr::map(rows$config_id, function(id) {
    f <- list.files(file.path("..", "data", id),
                    pattern = "\\.parquet$", full.names = TRUE)
    duckplyr::read_parquet_duckdb(f) |>
      collect() |>
      mutate(split = rows$split[match(id, rows$config_id)])
  }) |> bind_rows()
}

agreement_table <- function(size_val) {
  d <- read_all(size_val)
  if (!nrow(d)) {
    return(tibble::tibble(metric = character(), max_diff = double()))
  }
  key <- c("sim", "stream", "ci_method", "nboot", "proportion", "dist")
  pivots <- c("est", "lcl", "ucl")
  ## per-split tibble keyed by run config
  pivot <- d |>
    select(all_of(c(key, pivots, "split"))) |>
    pivot_wider(names_from = split, values_from = all_of(pivots))
  splits_present <- c("split0", "split1", "split2", "split3")
  splits_present <- intersect(splits_present, unique(d$split))
  if (length(splits_present) < 2) {
    return(tibble::tibble(metric = pivots, max_diff = NA_real_))
  }
  ref <- splits_present[1]
  others <- setdiff(splits_present, ref)
  purrr::map_dfr(pivots, function(m) {
    rcol <- paste0(m, "_", ref)
    diffs <- purrr::map_dbl(others, function(s) {
      ocol <- paste0(m, "_", s)
      if (!all(c(rcol, ocol) %in% names(pivot))) return(NA_real_)
      max(abs(pivot[[rcol]] - pivot[[ocol]]), na.rm = TRUE)
    })
    tibble::tibble(
      size     = size_val,
      metric   = m,
      ref      = ref,
      other    = others,
      max_abs_diff = diffs
    )
  })
}

agreement <- purrr::map_dfr(c("small", "medium", "large"), agreement_table)
agreement |> knitr::kable()
```

Zero (or floating-point-zero, e.g. `< 1e-10`) means the splits agree
exactly on the point estimate and CI bounds.

## Samples list-column

```{r}
samples_check <- function(size_val) {
  d <- read_all(size_val) |>
    mutate(samp_len = lengths(samples))
  d |>
    group_by(split, nboot, ci_method) |>
    summarise(
      n          = n(),
      uniq_lens  = paste(sort(unique(samp_len)), collapse = ","),
      .groups    = "drop"
    ) |>
    mutate(size = size_val) |>
    select(size, split, nboot, ci_method, n, uniq_lens)
}

purrr::map_dfr(c("small", "medium", "large"), samples_check) |>
  knitr::kable()
```

`uniq_lens` is the set of sample-vector lengths observed at each
`(nboot, ci_method)`. The samples vector either equals `nboot` (most CI
methods) or `nboot × #dists` (the multi methods); we expect this to be
identical across splits.

## Summary

If all three checks above are clean — row counts match, split-vs-split
estimate diff is at floating-point zero, and samples lengths are
consistent — then the split granularity is **purely an
infrastructure choice**: the answer doesn't depend on how the work was
chopped up.
