---
title: "Output correctness — `ssdsims` targets experiments"
format:
  gfm:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(duckplyr)
  library(readr)
  library(purrr)
})

summary_csv <- "../results/summary.csv"
stopifnot(file.exists(summary_csv))
sm <- readr::read_csv(summary_csv, show_col_types = FALSE)
```

The three split configs ran the same atomic fit grid; each wrote one
parquet file per `fit_dists_seed()` call to `data/<config_id>/fit-*.parquet`.
Branches differ in how many fits they bundle, but the *combined*
output should be identical across splits — only the packaging differs.

## Per-config row counts

```{r}
read_config <- function(id) {
  glob <- file.path("..", "data", id, "*.parquet")
  duckplyr::read_parquet_duckdb(glob)
}

counts <- sm |>
  rowwise() |>
  mutate(
    n_rows = read_config(config_id) |>
      summarise(n = n()) |> collect() |> pull(n)
  ) |>
  ungroup() |>
  select(split, n_branches, n_atomic_fits, n_hc_per_fit, n_rows)

counts |> knitr::kable()
```

Each atomic fit task runs `n_hc_per_fit` HC calls. With
`est_method = "multi"`, every `(nrow, sim, ci_method, nboot,
proportion, dist)` tuple contributes one row to the output:

```{r}
counts |>
  mutate(
    expected = n_atomic_fits * n_hc_per_fit,
    delta    = n_rows - expected
  ) |>
  select(split, n_rows, expected, delta) |>
  knitr::kable()
```

`delta` may be non-zero if some distributions don't yield rows for
particular configurations (the HC table includes one row per fitted
distribution × HC combo). It should be constant across splits.

If `delta` is zero everywhere, splitting is row-count-preserving.

## Cross-split agreement

For every `(sim, nrow, ci_method, nboot, proportion)` key the
`est`/`lcl`/`ucl` columns should match across splits — seeds are
deterministic, and the work per atomic unit doesn't depend on how
it's batched.

```{r}
read_all <- function() {
  purrr::map(sm$config_id, function(id) {
    read_config(id) |>
      collect() |>
      mutate(split = sm$split[match(id, sm$config_id)])
  }) |> bind_rows()
}

d <- read_all()
key <- c("sim", "stream", "nrow", "ci_method", "nboot", "proportion", "dist")
pivots <- c("est", "lcl", "ucl")

agreement <- if (length(unique(d$split)) < 2) {
  tibble::tibble(metric = pivots, max_abs_diff = NA_real_)
} else {
  pivot <- d |>
    select(all_of(c(key, pivots, "split"))) |>
    pivot_wider(names_from = split, values_from = all_of(pivots))
  splits_present <- intersect(
    c("by_nrow", "by_sim", "atomic"),
    unique(d$split)
  )
  ref <- splits_present[1]
  others <- setdiff(splits_present, ref)
  purrr::map_dfr(pivots, function(m) {
    rcol <- paste0(m, "_", ref)
    purrr::map_dfr(others, function(s) {
      ocol <- paste0(m, "_", s)
      if (!all(c(rcol, ocol) %in% names(pivot))) {
        return(tibble::tibble(metric = m, ref = ref, other = s,
                              max_abs_diff = NA_real_))
      }
      tibble::tibble(
        metric       = m,
        ref          = ref,
        other        = s,
        max_abs_diff = max(abs(pivot[[rcol]] - pivot[[ocol]]), na.rm = TRUE)
      )
    })
  })
}

agreement |> knitr::kable()
```

Zero (or floating-point-zero, e.g. `< 1e-10`) means the splits agree
exactly on the point estimate and CI bounds.

## Samples list-column

```{r}
d |>
  mutate(samp_len = lengths(samples)) |>
  group_by(split, nboot, ci_method) |>
  summarise(
    n         = n(),
    uniq_lens = paste(sort(unique(samp_len)), collapse = ","),
    .groups   = "drop"
  ) |>
  knitr::kable()
```

`uniq_lens` is the set of sample-vector lengths observed at each
`(nboot, ci_method)`. The samples vector either equals `nboot`
(most CI methods) or `nboot × #dists` (multi methods); we expect this
to be identical across splits.

## Summary

If row counts match, split-vs-split estimate diff is floating-point
zero, and samples lengths are consistent, then split granularity is
**purely an infrastructure choice**: the answer doesn't depend on
how the work was chopped up.
