---
title: "Split granularity vs. wall time — `ssdsims` targets experiments"
format:
  gfm:
    toc: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(scales)

summary_csv <- "../results/summary.csv"
meta_csv    <- "../results/per_target.csv"

stopifnot(file.exists(summary_csv), file.exists(meta_csv))

sm <- readr::read_csv(summary_csv, show_col_types = FALSE) |>
  mutate(
    size  = factor(size,  levels = c("small", "medium", "large")),
    split = factor(split, levels = c("split0", "split1", "split2", "split3"))
  )

meta <- readr::read_csv(meta_csv, show_col_types = FALSE) |>
  filter(type == "branch", parent == "branch_results") |>
  mutate(
    size  = factor(size,  levels = c("small", "medium", "large")),
    split = factor(split, levels = c("split0", "split1", "split2", "split3"))
  )

cores <- parallel::detectCores()
```

This pipeline branches a bootstrap simulation (scenario 1 of
`scripts/example.R`) into dynamic targets, then runs them via a
`crew::crew_controller_local(workers = `r cores`)`. We sweep over
**three problem sizes** and **four split granularities**, then ask:

- Where does the wall time go as we make work units smaller?
- What's the per-branch time distribution at each split?
- Is there a "sweet spot" for batch size?

## Configurations

```{r}
sm |>
  select(size, split, split_axes, n_atomic_units, n_branches,
         work_per_branch = n_atomic_units) |>
  mutate(work_per_branch = round(work_per_branch / n_branches, 1)) |>
  knitr::kable()
```

`n_atomic_units` is the total number of distinct `(nrow, ci_method, nboot,
proportion)` tuples; `n_branches` is how many of those tuples each become
their own dynamic target. `work_per_branch` is the mean fan-in.

## Total wall time

```{r}
ggplot(sm, aes(split, wall_secs, group = size, colour = size)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text(aes(label = paste0(round(wall_secs), "s")),
            vjust = -0.8, size = 3) +
  scale_y_continuous(labels = label_number(suffix = "s")) +
  labs(
    x = "Split granularity (coarse → fine)",
    y = paste0("Wall time on ", cores, " cores"),
    colour = "Problem size",
    title = "Wall time across split granularity"
  ) +
  theme_minimal()
```

## Parallel efficiency

```{r}
sm <- sm |>
  mutate(
    cpu_time      = branch_secs_total,
    parallel_eff  = cpu_time / (wall_secs * cores)
  )

sm |>
  select(size, split, n_branches, cpu_time, wall_secs, parallel_eff) |>
  mutate(
    cpu_time      = round(cpu_time, 1),
    wall_secs     = round(wall_secs, 1),
    parallel_eff  = scales::percent(parallel_eff, accuracy = 1)
  ) |>
  knitr::kable()
```

```{r}
ggplot(sm, aes(split, parallel_eff, group = size, colour = size)) +
  geom_line() +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Split granularity (coarse → fine)",
    y = paste0("Parallel efficiency (cpu / wall / ", cores, ")"),
    colour = "Problem size",
    title = "How much of the work actually parallelises"
  ) +
  theme_minimal()
```

## Per-branch time distribution

```{r}
ggplot(meta, aes(split, seconds, colour = size)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_y_log10() +
  facet_wrap(~size, scales = "free_y") +
  labs(
    x = "Split", y = "Per-branch seconds (log)",
    title = "Per-branch wall time distribution"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
meta |>
  group_by(size, split) |>
  summarise(
    n       = n(),
    min_s   = round(min(seconds), 2),
    p50_s   = round(median(seconds), 2),
    mean_s  = round(mean(seconds), 2),
    p90_s   = round(quantile(seconds, 0.9), 2),
    max_s   = round(max(seconds), 2),
    .groups = "drop"
  ) |>
  knitr::kable()
```

## Storage cost

```{r}
sm |>
  select(size, split, n_branches, parquet_count,
         parquet_KB = parquet_bytes) |>
  mutate(parquet_KB = round(parquet_KB / 1024, 1)) |>
  knitr::kable()
```

```{r}
ggplot(sm, aes(n_branches, parquet_bytes / 1024, colour = size)) +
  geom_point(size = 3) +
  geom_line(aes(group = size)) +
  scale_x_log10() +
  scale_y_continuous(labels = label_number(suffix = " KB")) +
  labs(
    x = "Number of branches (log)",
    y = "Total Parquet bytes",
    colour = "Problem size",
    title = "Output storage vs. number of branches"
  ) +
  theme_minimal()
```

## Picking a split

The trade-off in one sentence: **enough branches to keep the workers
saturated, big enough work units that crew dispatch / parquet-write
overhead is amortised, small enough that interruption costs at most one
unit.**

Concretely from the data above:

- `split0` (branch by `nrow` only) gives the **smallest wall time and the
  highest parallel efficiency**, but every branch is a 5–20 minute
  bootstrap — interrupting at minute 15 of 20 wastes 15 minutes.
- `split3` (atomic, one branch per `(nrow, ci_method, nboot, proportion)`)
  gives ~`r round(median(meta$seconds[meta$split == "split3"]), 1)`-second
  branches — interruption-safe to the second — but pays for it with
  noticeable crew dispatch overhead, lower parallel efficiency, and
  hundreds of tiny Parquet files.
- **Recommended sweet spot:** `split1` (branch by `nrow × ci_method`) or
  `split2` (add `nboot`) — branch wall time on the order of a minute,
  parallel efficiency still close to 1.0, branch count proportional to
  problem size rather than to atomic-unit count.

A rule of thumb that holds across sizes: **target branches of ~30–120 s**
on a 4-core machine; on HPC, scale up by ~`workers / 4` so per-branch time
still dwarfs scheduler latency (typically 5–15 s on SLURM).

## Caveats

- Local crew with `r cores` workers; HPC numbers (mirai/clustermq/SLURM)
  will differ. Scheduler latency adds a per-branch floor that punishes
  fine splits more on HPC than locally.
- `nsim = 2` for every config — branch time variance underestimates what
  you'd see with real sample sizes.
- Bootstrap cost grows roughly linearly with `nboot`; the largest `nboot`
  in any config is 100, capped to keep total experiment wall time bounded.
