---
title: "Split granularity vs. wall time — `ssdsims` targets experiments"
format:
  gfm:
    toc: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(readr)
library(scales)

summary_csv <- "../results/summary.csv"
meta_csv    <- "../results/per_target.csv"
stopifnot(file.exists(summary_csv), file.exists(meta_csv))

sm <- readr::read_csv(summary_csv, show_col_types = FALSE) |>
  mutate(split = factor(split, levels = c("by_nrow", "by_sim", "atomic")))

meta <- readr::read_csv(meta_csv, show_col_types = FALSE) |>
  filter(type == "branch", parent == "branch_results") |>
  mutate(split = factor(split, levels = levels(sm$split)))

n_workers <- unique(sm$n_workers)
stopifnot(length(n_workers) == 1)
```

Each **atomic task** is one `fit_dists_seed()` call — one parquet file
per task containing the full HC sweep downstream of that fit. The
harness builds the same atomic grid for each split and lets the choice
of **split granularity** show up as a `dplyr::group_by()` over the
fit-grid columns. Dynamic branching (`iteration = "group"`) iterates
over groups, so each branch (= "job") bundles one or more atomic
tasks.

Compute target: **`r n_workers` `crew_controller_local()` workers**.

## Configurations

```{r}
sm |>
  select(split, split_axes, n_atomic_fits, n_branches, n_fits_per_branch) |>
  knitr::kable()
```

`n_atomic_fits` is identical across rows (the grid doesn't change);
`n_branches` is the number of groups produced by the split, and
`n_fits_per_branch` shows how the work is packed.

## Total wall time

```{r}
ggplot(sm, aes(split, wall_secs)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(wall_secs, 1), "s")), vjust = -0.4) +
  scale_y_continuous(labels = label_number(suffix = "s"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    x = "Split granularity (coarse → fine)",
    y = paste0("Wall time on ", n_workers, " workers"),
    title = "Wall time across split granularity"
  ) +
  theme_minimal()
```

## Parallel efficiency

```{r}
sm <- sm |>
  mutate(parallel_eff = branch_secs_total / (wall_secs * n_workers))

sm |>
  select(split, n_branches, branch_secs_total, wall_secs, parallel_eff) |>
  mutate(
    branch_secs_total = round(branch_secs_total, 1),
    wall_secs         = round(wall_secs, 1),
    parallel_eff      = scales::percent(parallel_eff, accuracy = 1)
  ) |>
  knitr::kable()
```

```{r}
ggplot(sm, aes(split, parallel_eff, group = 1)) +
  geom_line(colour = "steelblue") +
  geom_point(size = 3, colour = "steelblue") +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, NA)) +
  labs(
    x = "Split granularity",
    y = paste0("Parallel efficiency (∑ branch secs / wall / ", n_workers, ")"),
    title = "How much of the work actually parallelises"
  ) +
  theme_minimal()
```

## Per-branch time distribution

```{r}
ggplot(meta, aes(split, seconds)) +
  geom_boxplot(outlier.size = 1, fill = "steelblue", alpha = 0.4) +
  geom_jitter(width = 0.1, height = 0, size = 1, alpha = 0.6) +
  labs(
    x = "Split", y = "Per-branch seconds",
    title = "Per-branch wall time distribution"
  ) +
  theme_minimal()
```

```{r}
meta |>
  group_by(split) |>
  summarise(
    n      = n(),
    min_s  = round(min(seconds), 2),
    p50_s  = round(stats::median(seconds), 2),
    mean_s = round(mean(seconds), 2),
    p90_s  = round(stats::quantile(seconds, 0.9), 2),
    max_s  = round(max(seconds), 2),
    .groups = "drop"
  ) |>
  knitr::kable()
```

## Storage cost

```{r}
sm |>
  select(split, n_branches, parquet_count, parquet_KB = parquet_bytes) |>
  mutate(parquet_KB = round(parquet_KB / 1024, 1)) |>
  knitr::kable()
```

## Picking a split

The trade-off in one sentence: **enough branches to keep the workers
saturated, big enough work units that crew dispatch / parquet-write
overhead is amortised, small enough that interruption costs at most one
unit.**

Reading across the three splits:

- **`by_nrow`** bundles all sims of one nrow into one job. Few, fat
  branches; high parallel efficiency on workers ≤ n_branches, but
  scheduler can't load-balance within an nrow.
- **`by_sim`** bundles one sim across all nrows. Different work
  profile per branch (one cheap small-nrow fit + one expensive
  large-nrow fit), still few branches.
- **`atomic`** runs each `(nrow, sim)` fit in its own branch.
  Maximum parallelism and finest interruption granularity at the
  cost of per-branch dispatch + parquet-write overhead paid `n` times.

The right split depends on the (fit cost vs. dispatch overhead) ratio
on your compute target. On HPC the overhead floor is much higher (≥ 5s
scheduler latency) so coarser splits win; on local crew the overhead
is small enough that atomic is often viable.

## Reproducing this at scale

The baseline grid is set up to finish in ~1 minute on 2 CPUs. To
explore the trade-off at production scale, bump knobs in
`configs.R` — bigger `nboot`, more `ci_method`s, more `nrow_levels`.
Each knob roughly multiplies branch count or per-branch cost in an
obvious way; the wall-time-vs-split shape stays the same, the
constants get bigger.
