---
title: "Running on a SLURM Cluster"
vignette: >
  %\VignetteIndexEntry{Running on a SLURM Cluster}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
# This guide describes a SLURM run; the cluster code chunks are not evaluated
# here (no scheduler when the vignette builds). The one live chunk lists the
# shipped template, so it only needs ssdsims itself.
evaluate <- requireNamespace("ssdsims", quietly = TRUE)
knitr::opts_chunk$set(eval = evaluate)
```

```{r}
#| label: library
library(ssdsims)
```

The ["Running a Sharded Pipeline"](sharded-pipeline.html) vignette runs a
scenario's shards through a `targets` pipeline on a single machine. This vignette
takes that **same pipeline to a SLURM cluster** — and the only thing that
changes is the `crew` controller (`TARGETS-DESIGN.md` §4). The per-task
`sample`/`fit`/`hc` results are byte-identical to the local run; the cluster is
just a faster backend.

You do **not** need to know `crew` or `targets` to follow this guide — only your
cluster's own (non-R) usage instructions. It is four steps: map your site's SLURM
instructions onto the controller, confirm the mapping with a preflight, run a
minimal first job, then swap in your own scenario. This guide targets **SLURM**;
`crew.cluster` also supports SGE, PBS/TORQUE, and LSF — see
[Targeting a non-SLURM cluster](#targeting-a-non-slurm-cluster-untested) below.

## The `cluster` template

The package ships an editable `cluster` template (a sibling of `small` and
`large`). Three ingredients assemble into the run, and **only B is
cluster-specific**:

| Ingredient | What | Where |
|---|---|---|
| **A — shape** | the per-shard target graph | `ssd_scenario_targets()` in `_targets.R` (reused verbatim) |
| **B — backend** | the SLURM `crew` controller (queue, modules, scratch, workers, walltime) | the **one editable block** in `controller.R` |
| **C — content** | the scenario (seed, datasets, grids) | the inline `scenario` block in `_targets.R` |

It is just four files — separating only what cannot be inlined:

```{r}
#| label: template-files
list.files(system.file("targets-templates", "cluster", package = "ssdsims"))
```

`controller.R` (the **one editable controller block**, sourced by both the
pipeline and the preflight), `_targets.R` (the clean pipeline: controller + the
inline scenario + factory), `preflight.R` (the standalone
connectivity/prerequisite check, kept out of the pipeline), and `run.R` (the
driver: preflight then `tar_make()`). Copy them to your project root:

```{r}
#| label: get-template
#| eval: false
dir <- system.file("targets-templates", "cluster", package = "ssdsims")
file.copy(list.files(dir, full.names = TRUE, all.files = TRUE, no.. = TRUE), ".")
```

## Step 1 — Map your site's SLURM instructions to the controller

Your real starting point is your site's documentation: how to log in, submit an
`sbatch` job, which partition/queue and account/allocation to use, the `module`
system that provides R, the scratch filesystem, and the walltime/core limits.
None of that mentions R, `crew`, or `targets`. This table bridges it to the
**one editable controller block** in `controller.R`
(`crew.cluster::crew_controller_slurm()` /
`crew.cluster::crew_options_slurm()`):

| Your site says… | …goes here |
|---|---|
| log in to the **login/submit node** | run `run.R` from there (where `sbatch` is on `PATH`) |
| submit with **`sbatch`** | nothing to set — `crew.cluster` calls `sbatch` for you |
| **partition / queue** name | `crew_options_slurm(partition = "...")` |
| **account / allocation** code | a `"#SBATCH --account=..."` line in `script_lines` |
| **`module load R/4.x`** (and any deps) | a `"module load R/4.x"` line in `script_lines` |
| **scratch filesystem** path | an `"export TMPDIR=/scratch/$USER/..."` line in `script_lines` |
| **walltime** limit (per job) | `crew_options_slurm(time_minutes = ...)` |
| **cores** per task | `crew_options_slurm(cpus_per_task = ...)` |
| **memory** per cpu | `crew_options_slurm(memory_gigabytes_per_cpu = ...)` |
| how many jobs you may run at once | `crew_controller_slurm(workers = ...)` |

So a site whose instructions read *"`sbatch` to the `short` queue, `module load
R/4.3`, account `abc123`, scratch under `/scratch/$USER`, 60-minute walltime"*
becomes:

```{r}
#| label: controller
#| eval: false
controller <- crew.cluster::crew_controller_slurm(
  workers = 4L, # max concurrent SLURM jobs
  seconds_idle = 60, # let an idle worker release its allocation
  options_cluster = crew.cluster::crew_options_slurm(
    partition = "short",
    time_minutes = 60,
    cpus_per_task = 1,
    memory_gigabytes_per_cpu = 4,
    script_lines = c(
      "#SBATCH --account=abc123",
      "module load R/4.3", # exposes R + the ssdsims binary library (below)
      "export TMPDIR=/scratch/$USER/ssdsims" # writable scratch for tempdir()
    )
  )
)
```

Anything not covered by a named argument (a constraint, a QOS, a `--gres`) goes
in `script_lines` as a literal `#SBATCH` line — `script_lines` is injected
verbatim into the generated `sbatch` script. Also set `expected_r_version` (in
`controller.R`) to the major.minor your `module load` provides, so the preflight
(step 2) can confirm it.

### Backend prerequisite — the worker install path

A SLURM worker is a fresh R process on a compute node. It must be able to
`library(ssdsims)` **without compiling** the dependency tree from source (compute
nodes are often offline or slow to build). The crew labs pinned this down with a
**ManyLinux binary install path**: install ssdsims and its dependencies as
pre-built Linux binaries (e.g. from the Posit Public Package Manager binary
repository for your distro) into the library your `module load R/...` exposes, so
the worker resolves everything at load time. Put the `module load` that exposes
that library in `script_lines`. The preflight in step 2 is the **login-node
prerequisite checker**: it fails loudly if a worker cannot load ssdsims, so you
fix the install/module path before launching the study.

## Step 2 — Confirm the mapping with the preflight

`preflight.R` is a standalone check, kept out of the main pipeline. Run it on its
own:

```{r}
#| label: preflight
#| eval: false
source("preflight.R") # or, from a shell:  Rscript preflight.R
```

It submits **one** SLURM job through your controller and, inside that job, checks
the things that actually break a real run:

1. **R resolves** at `expected_r_version` — else fix the `module load` line;
2. **`library(ssdsims)` loads** — else fix the worker install/module path (the
   ManyLinux binary path above);
3. **`tempdir()` is writable** — else fix the scratch `export TMPDIR=...` line.

A green preflight prints a small witness (the worker's R version + node id) and
means your controller block matches the site. A failed preflight **aborts before
the pipeline runs** and names exactly which check failed, with the fix:

- *no job dispatched* → controller / queue / account wiring;
- *R or ssdsims missing on the worker* → `module load` / install path;
- *scratch not writable* → the `TMPDIR` / scratch path.

So you debug the cluster wiring, never an obscure scenario error. `run.R` runs
this preflight for you before `tar_make()`, so a broken cluster never reaches the
expensive scenario shards.

## Step 3 — Run a minimal first job

For your first job, make the scenario tiny and cheap so you are testing the
**wiring**, not waiting on a big study. Temporarily shrink the `scenario` block
near the top of `_targets.R` to a single cheap cell:

```{r}
#| label: minimal-scenario
#| eval: false
scenario <- ssd_define_scenario(
  ssddata::ccme_boron,
  nsim = 2L,
  nrow = 5L,
  seed = 42L
)
```

Then run the pipeline from the login node:

```{r}
#| label: run
#| eval: false
source("run.R") # or, from a shell:  Rscript run.R
```

`run.R` runs the preflight first, then dispatches the scenario's shards across
SLURM jobs (independent shards run concurrently), writes one Parquet per shard
under `results/layout=<hash>/<step>/...`, unions them into
`results/layout=<hash>/summary.parquet`, and prints a peek at the estimates. That
is your first running cluster job, end to end.

::: {.callout-note}
**No cluster handy?** `run.R` guards: if `crew.cluster` is not installed or
`sbatch` is not on `PATH`, it aborts with a clear message naming the missing
prerequisite. To run the **same study off a cluster** (no scheduler), use the
`large` template — it builds the identical pipeline (same factory + scenario)
under a `crew::crew_controller_local()` controller, and its `run-serial.R`
asserts the results are byte-identical.
:::

## Step 4 — Swap in your own scenario

Once the minimal job succeeds, expand the `scenario` block in `_targets.R` back
to the full sweep the template ships (or to your own study; see
`large/scenario.R`). **Leave `controller.R` and the `ssd_scenario_targets()` call
unchanged** — the scenario and the factory call are scheduler-independent. Run
`run.R` again.

## Targeting a non-SLURM cluster (untested)

This template and guide target **SLURM**, but `crew.cluster` also ships
controllers for **SGE** (Sun Grid Engine / Son of Grid Engine), **PBS/TORQUE**,
and **LSF**, so the same pipeline can in principle run on those schedulers too.
ssdsims has only been exercised on SLURM — treat the others as **supported by
`crew.cluster` but untested here**, and lean on the preflight (step 2) to catch
wiring problems early.

Only **ingredient B** changes: in `controller.R`, swap the controller
constructor and its options function. The factory (A), the scenario (C), and the
whole preflight stay exactly the same.

| Scheduler | Controller | Options |
|---|---|---|
| SLURM | `crew_controller_slurm()` | `crew_options_slurm()` |
| SGE | `crew_controller_sge()` | `crew_options_sge()` |
| PBS / TORQUE | `crew_controller_pbs()` | `crew_options_pbs()` |
| LSF | `crew_controller_lsf()` | `crew_options_lsf()` |

The **cluster-independent parts carry over unchanged**:

- **`script_lines`** exists on every options function and works the same way —
  put your `module load R/...`, your account/project directive, and your scratch
  `export TMPDIR=...` there (these are shell / `#SCHEDULER` directives, not R).
  So the worker-prerequisite mapping (the ManyLinux binary path) and the step-2
  preflight are identical on every scheduler.
- `workers` and `seconds_idle` are `crew` settings on the controller, not
  scheduler-specific.

The **named resource arguments differ** by scheduler; anything without a named
argument goes in `script_lines` as a literal directive (`#$` for SGE, `#PBS` for
PBS/TORQUE, `#BSUB` for LSF):

| Resource | SLURM | SGE | PBS / TORQUE | LSF |
|---|---|---|---|---|
| cores per worker | `cpus_per_task` | `cores` | `cores` | `cores` |
| memory | `memory_gigabytes_per_cpu` | `memory_gigabytes_limit` | `memory_gigabytes_required` | `memory_gigabytes_limit` |
| walltime | `time_minutes` | `script_lines` (`#$ -l h_rt=...`) | `walltime_hours` | `script_lines` (`#BSUB -W ...`) |
| queue / partition | `partition` | `script_lines` (`#$ -q ...`) | `script_lines` (`#PBS -q ...`) | `script_lines` (`#BSUB -q ...`) |

For example, the SLURM controller from step 1 becomes, on SGE:

```{r}
#| label: controller-sge
#| eval: false
controller <- crew.cluster::crew_controller_sge(
  workers = 4L,
  seconds_idle = 60,
  options_cluster = crew.cluster::crew_options_sge(
    cores = 1,
    memory_gigabytes_limit = 4,
    script_lines = c(
      "#$ -q short.q", # queue (SGE has no `partition` argument)
      "#$ -l h_rt=01:00:00", # walltime
      "module load R/4.3",
      "export TMPDIR=/scratch/$USER/ssdsims"
    )
  )
)
```

See `?crew.cluster::crew_controller_sge` (and the `_pbs` / `_lsf` equivalents)
for the exact arguments. Everything else in this guide — the preflight, the
minimal first job, swapping in your scenario — is unchanged.

## Shard ↔ SLURM-job packing

Shards are the **unit of parallelism**: independent shard targets run
concurrently across your SLURM jobs. How many shard targets ride in one job is a
`crew` configuration knob (`workers` caps concurrent jobs; `tasks_max` on the
controller packs several shards into one worker's lifetime — "many-to-one";
`tasks_max = 1L` gives one shard per job — "one-to-one"). It is documented in
`controller.R`, not hard-coded as 1:1.

## See also

- The template's own `README.md` (the same guide, alongside the files) and
  `TARGETS-DESIGN.md` §4 (from local to a cluster), §11 (open questions), §12
  (`cluster-pipeline`).
- [`?ssd_scenario_targets`](../reference/ssd_scenario_targets.html),
  `?crew.cluster::crew_controller_slurm`.
- The ["Running a Sharded Pipeline"](sharded-pipeline.html) vignette — the
  `small`/`large`/`cluster` template trio.
