---
title: "Analysing a Run's Observed Compute Cost"
vignette: >
  %\VignetteIndexEntry{Analysing a Run's Observed Compute Cost}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
# The vignette runs a small scenario through the single-core shard runner and
# reads its timings back, so it needs the fitting deps and duckplyr. Skip
# evaluation gracefully on a minimal runner rather than failing the build.
evaluate <- requireNamespace("ssddata", quietly = TRUE) &&
  requireNamespace("ssdtools", quietly = TRUE) &&
  requireNamespace("duckplyr", quietly = TRUE)
knitr::opts_chunk$set(eval = evaluate)
```

```{r}
#| label: library
library(ssdsims)
```

## Prediction, then measurement

[`ssd_estimate_cost()`](../reference/ssd_estimate_cost.html) predicts a
scenario's compute cost *before* a run (see
[`vignette("cost-estimation")`](cost-estimation.html)), from a cost model
calibrated by a synthetic micro-benchmark. Once a run has happened it leaves
**ground truth** behind: every `fit` and `hc` task's
body is bracketed with a UTC `.start`/`.end` and tagged with the `.host` CPU, and
those land as columns of the shard Parquets themselves. `ssd_analyse_cost()` reads
them back and attributes the *observed* compute to the same `ci_method` × `nboot`
axes the estimate uses --- closing the loop from predicted to measured.

Like the estimator, the analysis functions are strictly **read-only**: they read
result Parquets (and, optionally, the run's `targets` store), and never run the
pipeline, fit a distribution, draw a random number, or write a file.

## A small worked run

Any run leaves timings behind --- the serial baseline, the single-core
[`ssd_run_scenario_shards()`](../reference/ssd_run_scenario_shards.html), or the
`targets` pipeline. Here we materialise a small scenario single-core so the
vignette is self-contained:

```{r}
#| label: run
data <- ssd_scenario_data(ssddata::ccme_boron)
scenario <- ssd_define_scenario(
  data,
  nsim = 2L,
  seed = 42L,
  nrow = c(6L, 10L),
  ci = TRUE,
  nboot = c(100L, 1000L),
  dists = ssd_distset(lnorm = "lnorm")
)

run <- ssd_run_scenario_shards(scenario)
```

## Analysing observed cost

`ssd_analyse_cost()` reads the run's `fit`/`hc` shard timings and reports the
observed **total** compute and the observed **longest single task**, with a
breakdown by `ci_method` × `nboot` keyed exactly like the estimate's:

```{r}
#| label: analyse
analysis <- ssd_analyse_cost(scenario, root = run$dir)
analysis
```

The observed **total is serial-equivalent compute** --- the sum of the per-task
durations --- *not* elapsed wall time. Under `n` parallel workers the wall time
is roughly `max(longest_task, total / n)`, so the longest single task is the
irreducible floor regardless of how many workers you add.

### The shard envelope

Per-task durations measure the *work*; they do not include the per-shard overhead
of reading the parent shard, writing the Parquet, and `crew` dispatch. Given the
run's `targets` store, `ssd_analyse_cost()` reads each shard target's wall
`seconds` from `tar_meta()` and reports the **envelope overhead** ---
`target seconds − Σ task durations` --- the number that informs `partition_by`
tuning:

```{r}
#| label: envelope
#| eval: false
# With a targets store (a pipeline run), add the per-shard envelope overhead.
ssd_analyse_cost(scenario, root = "results", store = "_targets")
```

## Comparing predicted against observed

`ssd_compare_cost()` places the prediction beside the observation and reports the
predicted/observed ratio for the total and the longest task --- a one-glance check
of how well the (machine-specific) cost model tracked reality:

```{r}
#| label: compare
ssd_compare_cost(scenario, root = run$dir)
```

A ratio far from `1` means the shipped default calibration does not match this
machine --- which is exactly what recalibration fixes.

## Recalibrating from the run

`ssd_calibrate_cost_from_run()` re-fits the per-task cost model from the run's
**measured** hc task durations (and a measured fit addend), returning the same
`ssdsims_cost_calibration` object
[`ssd_calibrate_cost()`](../reference/ssd_calibrate_cost.html) produces --- so it
drops straight back into `ssd_estimate_cost()`, now grounded in
real measurements from this scenario rather than a synthetic sweep:

```{r}
#| label: recalibrate
from_run <- ssd_calibrate_cost_from_run(scenario, root = run$dir)

# The run-derived calibration drives the estimator like any other.
ssd_estimate_cost(scenario, calibration = from_run)
```

Because the calibration is **architecture-specific**, `ssd_calibrate_cost_from_run()`
never silently pools timings from different CPUs: a run whose shards span more
than one `.host` (a cluster of mixed node types) requires an explicit `host =`,
or the function aborts listing the hosts it found. The result's provenance records
that it was derived from a run.

## Across a design

A design (`ssd_design()`; see
[`vignette("scenario-to-design")`](scenario-to-design.html)) runs several
scenarios as one pipeline. The same three functions accept a design and **roll the observed cost up
across its members**: `ssd_analyse_cost(design, root = ...)` returns a breakdown
with a leading `scenario` column and design totals, `ssd_compare_cost(design)`
places the design's summed prediction beside it, and
`ssd_calibrate_cost_from_run(design)` pools every member's measured durations into
one host-aware calibration.

```{r}
#| label: design
#| eval: false
design <- ssd_design(coarse = scenario, dense = other_scenario)

# Per-scenario observed breakdown + design totals (read from the run's results
# root; the combined <root>/summary.parquet is used as a one-read fast path when
# present, else each member is read from its seed group's shared shards).
ssd_analyse_cost(design, root = "results")

# Predicted (Σ members) vs observed, with ratios.
ssd_compare_cost(design, root = "results")

# One calibration pooled across all members (host-aware).
ssd_calibrate_cost_from_run(design, root = "results")
```

Because `ssd_design_targets()` writes one **shared** `seed=/layout=` tree per seed
group --- a cell shared by several members is computed once --- the design total is
**per-member accounting**: a shared cell counts toward each member that includes
it (it answers "what does each scenario cost"), so the design total can exceed the
design's once-built shared compute. As at the scenario level, recalibration never
silently pools timings across CPUs: a design spanning more than one `.host`
requires an explicit `host =`.

## The loop

Putting it together --- predict, run, measure, compare, recalibrate, predict
again:

```{r}
#| label: loop
#| eval: false
estimate <- ssd_estimate_cost(scenario)        # predict
run <- ssd_run_scenario_shards(scenario)        # run
ssd_compare_cost(scenario, root = run$dir)      # predicted vs observed
better <- ssd_calibrate_cost_from_run(scenario, root = run$dir) # recalibrate
ssd_estimate_cost(next_scenario, calibration = better)          # predict the next, from truth
```

Each larger scenario is then sized from the measured cost of the last, rather
than a synthetic micro-benchmark --- the estimate improves every time the loop
turns.

## See also

- [`vignette("cost-estimation")`](cost-estimation.html) --- the complement:
  *predict* a scenario's cost before running it.
- [`vignette("sharded-pipeline")`](sharded-pipeline.html) --- the shard runs
  whose per-task timings this reads back.
- [`ssd_analyse_cost()`](../reference/ssd_analyse_cost.html),
  [`ssd_compare_cost()`](../reference/ssd_compare_cost.html),
  [`ssd_calibrate_cost_from_run()`](../reference/ssd_calibrate_cost_from_run.html).
