---
title: "Defining a Scenario"
vignette: >
  %\VignetteIndexEntry{Defining a Scenario}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
# The vignette exercises the live API, so it needs the optional fitting
# dependencies. Skip evaluation gracefully if they are unavailable (e.g. on a
# minimal CI runner) rather than failing the build.
evaluate <- requireNamespace("ssddata", quietly = TRUE) &&
  requireNamespace("ssdtools", quietly = TRUE)
knitr::opts_chunk$set(eval = evaluate)
```

```{r}
#| label: library
library(ssdsims)
```

## Why a declarative scenario?

ssdsims runs a simulation study from a declarative description rather than
immediately, feeding a cluster-friendly [targets](https://docs.ropensci.org/targets/)
pipeline. The root of that pipeline is a **scenario**: a small, purely
declarative description of the study you want to run. It contains only the
scenario options --- a seed, the number of simulations, the sample sizes, the
dataset *names*, and the fit/hc argument grids. It draws **no** random numbers,
expands **no** tasks, writes **nothing**, and has **no** dependency on
`targets`.

Keeping the scenario declarative buys two things:

- It **serialises to a compact manifest** --- no data frames, no RNG state, no
  function bodies --- so it can be shipped to a cluster and stored alongside
  results.
- The set of work a pipeline expands to is a **pure function of the scenario**,
  so the same scenario always describes the same study.

This vignette walks the stages that work *today*, in order. It is intended to
grow: as roadmap features land (per-task seeding, shards, the `targets`
backend), new sections will document them here. See `TARGETS-DESIGN.md` for the
north-star design.

The four stages this vignette covers:

1. **Assemble** the data with `ssd_scenario_data()` (and, for generated data,
   `ssd_gen()`).
2. **Declare** the scenario with `ssd_define_scenario()`.
3. **Expand** it into per-step task tables with `ssd_scenario_tasks()`.
4. **Run** the baseline in-process loop with `ssd_run_scenario_baseline()`.

Stages 2--3 are side-effect-free; generation (`ssd_gen()`, in stage 1) and
stage 4 draw random numbers under scoped, seeded dqrng state.

## Stage 1: assemble the data

`ssd_scenario_data()` is the single entry point for dataset input. It validates each
data frame --- every dataset must carry a numeric `Conc` column, the species
sensitivity distribution convention --- and assembles them into a named
collection.

```{r}
#| label: ssd-data
data <- ssd_scenario_data(
  boron = ssddata::ccme_boron,
  cadmium = ssddata::ccme_cadmium
)
data
```

Names come from the argument names where you supply them; otherwise they are
derived from the argument expression by symbol capture (so a bare
`ssddata::ccme_boron` becomes `"ccme_boron"`). Names must be unique, and a
literal with no derivable name (e.g. a bare `data.frame(...)`) must be given an
explicit name.

### Generated datasets with `ssd_gen()`

Each dataset can also be *generated* rather than observed. `ssd_gen()` accepts
generator inputs ---
a generator function, a function-name string, or a `fitdists`/`tmbfit` object
(drawn from the matching `ssd_r<dist>` with the fit's estimates) --- and
materialises each, **once**, to a `Conc` tibble of `.n` rows. The result
composes with the data frames in `ssd_scenario_data()`, either as an unnamed
argument or spliced with `!!!` (the two are equivalent):

```{r}
#| label: ssd-gen
ssd_scenario_data(
  boron = ssddata::ccme_boron,
  ssd_gen(synth = ssdtools::ssd_rlnorm, .n = 30, .seed = 1L)
)
```

Both `.n` (the generated population size) and `.seed` are **required**, so
unsized or irreproducible generation is impossible. Each generator draws under
the dqrng `pcg64` backend on an independent stream keyed by its dataset *name*
(with `.seed` as the base seed), so one `.seed` reproducibly fans out across all
generators in a call; the global RNG state is untouched.

A materialised generator dataset is a *fixture the scenario resamples*: once
inside the collection it is an ordinary tibble, indistinguishable from a
data-frame dataset downstream (the generator code is its provenance).

## Stage 2: declare the scenario

`ssd_define_scenario()` is the constructor. Its required arguments are the
dataset input, `nsim` (the replicate count), and `seed` (the RNG root --- it has
no default because changing it fully re-roots every draw). Everything else is a
scenario option with a sensible default.

```{r}
#| label: define-scenario
scenario <- ssd_define_scenario(
  data,
  nsim = 3L,
  seed = 42L,
  nrow = c(5L, 10L),
  dists = ssd_distset(
    BCANZ = ssdtools::ssd_dists_bcanz(),
    lnorm = "lnorm"
  ),
  est_method = "multi",
  proportion = c(0.05, 0.2),
  ci = TRUE,
  nboot = c(10L, 50L),
  ci_method = "weighted_samples"
)
scenario
```

The `print()` method shows the declarative fields: the scalar `seed` and
`nsim`, the `nrow` sample sizes, the dataset names, and the two argument grids
(`fit` and `hc`). Note what is *not* shown --- the data frames themselves are
retained for the local runner but are not part of the declarative identity; the
cluster path carries only the names.

### Dataset input is the collection

Dataset input is accepted **only** as an `ssd_scenario_data()` collection ---
naming, validation, and (via `ssd_gen()`) generation all live there, so the
constructor itself stays inert and draws no random numbers:

```{r}
#| label: input-form
data <- ssd_scenario_data(ssddata::ccme_boron)
ssd_define_scenario(data, nsim = 2L, seed = 1L)
```

### `min_pmix` is referenced by name

The `min_pmix` scenario option is referenced **by name**: the name --- not the
function body --- is what enters the task identity and hashes, so the scenario's
identity stays stable under a recompile or a cosmetic edit. It is supplied as an
`ssd_pmix()` collection (the only accepted form --- a bare function, a plain
list, or a character vector of names is rejected, and no name-string is resolved
to a function). Names come from the `ssd_pmix()` argument names, or by symbol
capture for a bare `pkg::name` (mirroring `ssd_scenario_data()`). The default
`ssd_pmix(ssd_min_pmix = ssdtools::ssd_min_pmix)` is stored as `"ssd_min_pmix"`:

```{r}
#| label: min-pmix
scenario$fit$min_pmix
```

The single-argument *function* from the collection is additionally
**materialised on the scenario** at construction. It rides along for execution
and is retrieved by name with the `scenario_min_pmix()` accessor; datasets are
reached the same way with `scenario_dataset()`:

```{r}
#| label: accessors
identical(scenario_min_pmix(scenario, "ssd_min_pmix"), ssdtools::ssd_min_pmix)
head(scenario_dataset(scenario, "boron"), 3)
```

Because the *name* drives hashing while the *value* only rides along for
execution, two scenarios with the same `min_pmix` name but different function
bodies produce byte-identical task identities --- the split that lets a cluster
worker resolve `min_pmix` off the transported scenario with no shared
interactive environment.

### `dists` is a collection of distribution sets

`dists` is supplied as an `ssd_distset()` collection: one or more named
**distribution sets**, each a pool of distributions model-averaged together to
form one SSD. The constructor validates members (a subset of
`ssdtools::ssd_dists_all()`) and names (unique, filesystem-safe) up front:

```{r}
#| label: distset
ds <- ssd_distset(
  BCANZ = ssdtools::ssd_dists_bcanz(),
  Iwasaki = c("burrIII3", "gamma", "llogis", "lnorm", "weibull"),
  lnorm = "lnorm"
)
ds
```

The fit step fits the **union** of every set's members *once* --- the single
model-averaged superset every pool is a subset of --- and the hc step then
`subset()`s that one fit down to each pool's members and re-averages. So
declaring several pools costs **one** fit, not one per pool (the motivating
`ssdaveragerr` "iwasaki" pattern: fit a superset, then compare BCANZ, the
Iwasaki set, and single distributions carved from it):

```{r}
#| label: distset-union
scenario_iwasaki <- ssd_define_scenario(
  data,
  nsim = 3L,
  seed = 42L,
  dists = ds
)
# one union fit per fit task ...
scenario_iwasaki$fit$dists
# ... and the named pools carried for the hc step to subset by name
names(scenario_iwasaki$hc$distsets)
```

The set **name** --- not its members --- is what enters the task identity (the
`distset=<name>` path segment and the per-task primer), mirroring `min_pmix` and
datasets; the members ride on the scenario and are reached by name with the
`scenario_distset()` accessor:

```{r}
#| label: distset-accessor
scenario_distset(scenario_iwasaki, "Iwasaki")
```

`distset` is an **hc** cross-join axis, so the hc task table fans out over the
declared sets (the fit table is unchanged --- it fits the union). A single-set
collection (`ssd_distset(BCANZ = ...)`) has one `distset` value and so does not
multiply the table. Individual distributions still never fan out: an axis value
is always a whole averaging pool. A bare character vector or plain list is
rejected, pointing you to `ssd_distset()`.

### The `ci` flag

`ci` is a scalar flag, not a cross-join axis. The point estimate is computed
analytically from the fit and is identical whether `ci` is `TRUE` or `FALSE`,
so a `ci = TRUE` run is a strict superset of `ci = FALSE` (the same estimate,
plus the `se`/`lcl`/`ucl` columns). The choice is a scenario-wide either/or:
`ci = TRUE` for estimates plus bootstrap intervals (as in the scenario above),
or `ci = FALSE` for cheap, bootstrap-free point estimates only.

When `ci = FALSE`, the bootstrap-only scenario options (`nboot`, `ci_method`,
`parametric`) are meaningless, so passing any of them is an error --- set
`ci = TRUE` to enable the bootstrap, or omit the option:

```{r}
#| label: ci-false-error
#| error: true
data <- ssd_scenario_data(ssddata::ccme_boron)
ssd_define_scenario(data, nsim = 2L, seed = 1L, ci = FALSE, nboot = 1000)
```

## Stage 3: expand into task tables

`ssd_scenario_tasks()` expands the scenario into the three per-step task tables
--- `sample`, `fit`, and `hc` --- bundled into a task set. This is
still RNG-free: no draws happen here.

```{r}
#| label: tasks
tasks <- ssd_scenario_tasks(scenario)
tasks
```

The counts show how each step fans out from its parent. The key design choice
is that the single expensive **random draw** is its own `sample` task, keyed
only by `(dataset, sim, replace)`. The draw is sized by the scenario's fixed
`nrow_max` setting (`nrow_max` resampled rows under the default
`replace = TRUE`; capped at the dataset size --- the full permutation --- when
`replace = FALSE`); every `nrow` value is then a cheap,
RNG-free `head()` *sub-truncation* of that one draw, done inline at the `fit`
step. So `nrow` is an ordinary cross-join axis of the `fit` step and never
multiplies the underlying draw --- one draw is shared across all sample sizes,
and adding an `nrow` value (within the effective draw size) never changes it.

Each row carries a path-style `<step>_id` primary key (the Hive partition path)
plus its parent step's id as a foreign key, so dependencies are explicit and
joinable:

```{r}
#| label: sample-table
tasks$sample
```

The scalar `ci` is applied uniformly to every `hc` task (it is a setting read
from the scenario, not an axis or a table column), as is `est_method`. With
`ci = TRUE` the bootstrap axes (`nboot`, `ci_method`, `parametric`) fan out
fully; a `ci = FALSE` scenario would instead give exactly one `hc` row per fit
task, carrying `NA` for those bootstrap-only scenario options (every requested
`est_method` is summarised within each task from a single bootstrap, so it never
fans out into separate tasks).

```{r}
#| label: hc-table
tasks$hc
```

The per-step derivations are also available individually
(`ssd_scenario_sample_tasks()`, `ssd_scenario_fit_tasks()`,
`ssd_scenario_hc_tasks()`) when you only need one table.

## Stage 4: run the baseline loop

`ssd_run_scenario_baseline()` is the no-frills runner. It executes the three
task tables in dependency order, threading each task's result forward to its
children by the foreign-key id. It runs **in-process**, with no `targets`, no
shard grouping, and no Parquet I/O.

It **is reproducible without an external seed**: the runner opens one
`local_dqrng_backend()` scope and seeds each task exactly once from
`scenario$seed` plus a per-task primer (`task_primer()` over the task's
identity), so two runs of a scenario with a fixed `seed` yield identical
results, regardless of the order in which tasks run:

```{r}
#| label: run-baseline
out <- ssd_run_scenario_baseline(scenario)
names(out)
```

Each element is the corresponding task table augmented with a list column of
per-task results --- `sample` draws, `fits` objects, and `hc` tibbles (the
`fit` step truncates its sample inline before fitting):

```{r}
#| label: hc-out
out$hc
```

Unnest the hazard-concentration estimates back onto their task identities to
get a tidy results table. `proportion` and `est_method` are not task axes ---
they are passed whole to each `hc` call (one row per value, from a single
bootstrap), so every result already carries its own `proportion` and
`est_method` columns:

```{r}
#| label: unnest
hcs <- tidyr::unnest(
  out$hc[c("dataset", "sim", "nrow", "hc")],
  hc,
  names_sep = "_"
)
hcs[c("dataset", "sim", "nrow", "hc_est_method", "hc_proportion", "hc_est")]
```

## Next: shards and the targets pipeline

The baseline runner threads results in memory. The next layer materialises each
step as **Hive-partitioned Parquet shards** grouped by the scenario's
`partition_by` axes, and links steps by reading parent shards back --- the
storage hand-off the cluster `targets` pipeline uses. The
["Running a sharded pipeline"](sharded-pipeline.html) vignette covers that: the
single-core `ssd_run_scenario_shards()` (byte-identical to the baseline runner)
and the shipped `targets` template.

Still ahead on the roadmap (`TARGETS-DESIGN.md` §12): a per-scenario manifest,
cloud upload, shard-completeness assertions, and the crew/SLURM controller. As
each lands, the vignettes grow to cover it.
