---
title: "Generating and Assembling Datasets"
vignette: >
  %\VignetteIndexEntry{Generating and Assembling Datasets}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
# The vignette exercises the live API, so it needs the optional generator
# dependencies. Skip evaluation gracefully if they are unavailable (e.g. on a
# minimal CI runner) rather than failing the build.
evaluate <- requireNamespace("ssddata", quietly = TRUE) &&
  requireNamespace("ssdtools", quietly = TRUE)
knitr::opts_chunk$set(eval = evaluate)
```

```{r}
#| label: library
library(ssdsims)
```

Before a scenario can be declared with `ssd_define_scenario()`, its datasets
have to be *assembled*: named, validated, and (where a dataset is synthetic)
*generated*. That is the first stage of the workflow the
["Defining a Scenario"](defining-a-scenario.html) vignette walks through, and it
is handled by exactly two functions:

- `ssd_scenario_data()` --- the single entry point for dataset input. It
  validates each data frame and assembles them into a named collection.
- `ssd_gen()` --- the generation half. It materialises *generator*-style inputs
  (a distribution draw function, a fitted SSD, and so on) into the same kind of
  validated dataset, once and reproducibly.

This vignette documents both in detail: every argument, the accepted input
forms, the structure of what they return, and how they compose. Neither
function draws random numbers at scenario-definition time except `ssd_gen()`,
whose generation is deliberately separated out precisely so that
`ssd_define_scenario()` itself stays RNG-free.

## `ssd_gen()`: materialise generated datasets

`ssd_gen()` takes one or more *generators* and materialises each --- **once** ---
to a validated dataset: a tibble with a numeric `Conc` column of `.n` rows.
"Materialise once" is the key idea: the output is a fixed tibble of numbers, not
a rule that keeps drawing. Downstream it is an ordinary dataset that the
scenario resamples, indistinguishable from an observed one.

### The four generator kinds

Each argument in `...` is a generator, and there are four accepted kinds:

- a **function** taking the number of rows as its first argument (e.g.
  `ssdtools::ssd_rlnorm`);
- a **function-name string** (e.g. `"ssd_rlnorm"`), resolved as a bare name in
  the caller's environment and then the `ssdtools` namespace (never by parsing
  code); the string doubles as the dataset name;
- a **`tmbfit`** object --- one distribution of an `ssdtools::ssd_fit_dists()`
  fit --- drawn from the matching `ssd_r<dist>` function with the fit's
  estimates;
- a **`fitdists`** object --- the top-weighted distribution is selected and then
  drawn as for a `tmbfit`.

A data frame is **not** a generator; pass observed data straight to
`ssd_scenario_data()`. Anything that is neither a function, a single
function-name string, a `tmbfit`, nor a `fitdists` is rejected.

### Arguments

`ssd_gen(..., .n, .seed)` has just two options beyond the generators
themselves, and both are **required** --- by construction, generation can be
neither unsized nor unseeded:

| Argument | Meaning | Requirement |
|----------|---------|-------------|
| `...`    | One or more generators (the four kinds above), optionally named. | At least one. |
| `.n`     | The number of rows each generator materialises --- the generated *population* size. | Required; a scalar whole number greater than 0. |
| `.seed`  | The base seed for generation, independent of the scenario's own `seed`. | Required; a scalar whole number. |

The two options are dot-prefixed (`.n`, `.seed`) so that a generator you happen
to name `n =` or `seed =` is never partial-matched onto them and is never
absorbed into `...`.

Note the division of labour: `.n` is the size of the *generated population*,
while the *scenario's* `nrow`/`nrow_max` options control the resample and
truncation sizes drawn from it later. One `.n` applies to the whole call; to
generate populations of differing sizes, splice several `ssd_gen()` calls
together.

### A worked example

Give each generator a name and `ssd_gen()` returns a named collection of
materialised datasets:

```{r}
#| label: gen-basic
gen <- ssd_gen(synth = ssdtools::ssd_rlnorm, .n = 30, .seed = 1L)
gen
```

Each member is a validated `Conc` tibble of exactly `.n` rows:

```{r}
#| label: gen-member
gen$synth
nrow(gen$synth)
```

A single call can materialise several generators at once. Because each is keyed
by its own dataset **name** (see below), one `.seed` fans out across them,
giving independent draws rather than identical ones:

```{r}
#| label: gen-many
ssd_gen(
  a = ssdtools::ssd_rlnorm,
  b = ssdtools::ssd_rlnorm,
  .n = 5,
  .seed = 1L
)
```

### How generators are named

Names follow a fixed precedence, and must be unique across the call:

1. the **argument name**, where you supply one (`synth = ...`);
2. otherwise, for the function-name-string form, the **string itself**
   (`"ssd_rlnorm"` becomes the name `"ssd_rlnorm"`);
3. otherwise, the name is **derived from the argument expression** by symbol
   capture (a bare `ssdtools::ssd_rlnorm` becomes `"ssd_rlnorm"`).

An input with no derivable name --- an anonymous function literal, say --- must
be given an explicit name, or `ssd_gen()` aborts asking for one.

### Reproducibility and the name-keyed streams

Generation draws under a scoped dqrng `pcg64` backend
(`local_dqrng_backend()`). Each generator is seeded through
`local_dqrng_state()` with `.seed` as the base seed and its own dataset **name**
as the stream key (`task_primer()` over `list(dataset = name)`). Two
consequences follow:

- **One `.seed` reproducibly fans out** across every generator in the call, on
  independent, name-keyed streams --- which is why the `a`/`b` example above
  drew two different samples from a single `.seed`.
- **The global RNG state is left untouched**: `ssd_gen()` restores
  `.Random.seed` on return, so calling it never perturbs the surrounding
  session's random stream.

Re-running the same call therefore reproduces the same data exactly:

```{r}
#| label: gen-reproducible
identical(
  ssd_gen(synth = ssdtools::ssd_rlnorm, .n = 30, .seed = 1L),
  ssd_gen(synth = ssdtools::ssd_rlnorm, .n = 30, .seed = 1L)
)
```

As a guard on that reproducibility, each draw is bracketed by a dqrng-integrity
check: a generator that *escapes* the backend --- for instance by switching
`RNGkind()` and drawing from base R, which would not be reproducible under
`.seed` --- aborts. A generator that consumes no randomness passes. A generator
is also required to return a numeric vector of length `.n`; anything else is an
error.

### Generating from a fitted distribution

The `tmbfit` and `fitdists` kinds let you generate data from a distribution
*fitted to real data* rather than one you parameterise by hand. Fit an SSD, hand
the fit to `ssd_gen()`, and it draws from the fit's own estimates (for a
`fitdists`, from its top-weighted distribution):

```{r}
#| label: gen-fit
fit <- ssdtools::ssd_fit_dists(ssddata::ccme_boron, dists = "lnorm")
ssd_gen(from_fit = fit, .n = 20, .seed = 7L)
```

## `ssd_scenario_data()`: assemble and validate the collection

`ssd_scenario_data()` is the single entry point through which
`ssd_define_scenario()` takes dataset input. It collects one or more datasets
into a validated, named collection.

### Arguments and validation

`ssd_scenario_data(...)` takes a single `...`: one or more data frames
(optionally named) and/or `ssd_gen()` collections. Every **data frame** is
validated to carry a numeric `Conc` column --- the species sensitivity
distribution convention --- and is coerced to a tibble; any additional columns
are preserved. A data frame that is not a data frame, or that lacks a numeric
`Conc` column, aborts.

```{r}
#| label: data-basic
data <- ssd_scenario_data(
  boron = ssddata::ccme_boron,
  cadmium = ssddata::ccme_cadmium
)
data
```

Naming mirrors `ssd_gen()`: names come from the argument names where supplied,
otherwise they are derived from the argument expression by symbol capture (so a
bare `ssddata::ccme_boron` becomes `"ccme_boron"`). Names must be unique, and a
literal with no derivable name (a bare `data.frame(...)` call, say) must be
given an explicit name.

```{r}
#| label: data-symbol-capture
ssd_scenario_data(ssddata::ccme_boron)
```

### Relationship to `ssd_gen()`

`ssd_gen()` does not stand alone in the workflow: its output is designed to
**compose into** `ssd_scenario_data()`. `ssd_scenario_data()` does not call
`ssd_gen()`, nor the reverse --- generation and assembly stay separate, so the
assembler draws no random numbers. Instead, an `ssd_gen()` collection passed to
`ssd_scenario_data()` is *flattened in*: each already-validated member joins the
collection under its own name, alongside the observed data frames.

There are two equivalent ways to pass it, and an `ssd_gen()` collection must
always be supplied **unnamed** (its members already carry their own names):

```{r}
#| label: data-gen-compose
# as an unnamed argument (flattened in) ...
ssd_scenario_data(
  boron = ssddata::ccme_boron,
  ssd_gen(synth = ssdtools::ssd_rlnorm, .n = 30, .seed = 1L)
)
```

```{r}
#| label: data-gen-splice
# ... or spliced with `!!!` --- exactly equivalent
ssd_scenario_data(
  boron = ssddata::ccme_boron,
  !!!ssd_gen(synth = ssdtools::ssd_rlnorm, .n = 30, .seed = 1L)
)
```

### The returned collection

`ssd_scenario_data()` returns an `ssdsims_data` object: a named list of
validated tibbles. Once a generated dataset is inside the collection it is an
ordinary tibble, indistinguishable downstream from an observed one --- the
generator code is simply its provenance. This collection is what
`ssd_define_scenario()` accepts as its dataset input:

```{r}
#| label: data-into-scenario
data <- ssd_scenario_data(
  boron = ssddata::ccme_boron,
  ssd_gen(synth = ssdtools::ssd_rlnorm, .n = 30, .seed = 1L)
)
ssd_define_scenario(data, nsim = 2L, seed = 42L)
```

## A note on seeds

Two independent seeds sit either side of this stage, and it is worth keeping
them distinct:

- `ssd_gen()`'s `.seed` controls **generation** --- the one-off draw that
  materialises a synthetic population. It is fixed at assembly time and baked
  into the resulting tibble.
- `ssd_define_scenario()`'s `seed` controls the **scenario's** downstream
  draws --- the resampling and bootstrapping the pipeline performs on whatever
  datasets it is given, generated or observed.

Because a generated dataset is materialised once and then treated as ordinary
data, the scenario resamples it under `seed` exactly as it would an observed
dataset; `.seed` no longer plays any role past generation.

## See also

- `vignette("defining-a-scenario")` --- the full workflow this data-assembly
  stage feeds into.
- `ssd_scenario_data()`, `ssd_gen()`, `ssd_define_scenario()`.
