---
title: "From a Single Scenario to a Design"
vignette: >
  %\VignetteIndexEntry{From a Single Scenario to a Design}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
evaluate <- requireNamespace("ssddata", quietly = TRUE) &&
  requireNamespace("ssdtools", quietly = TRUE)
knitr::opts_chunk$set(eval = evaluate)
```

```{r}
#| label: library
library(ssdsims)
```

A single `ssd_scenario` is one **regular** grid --- a rectangular cross-join of
the axes. A real study is rarely rectangular: you want a coarse grid everywhere
*and* a dense refinement in one region (more `nrow` values, but only for one
dataset, say). Forcing that into one scenario computes a cross-product you never
wanted; running it as several separate pipelines recomputes the cells they share
and gives you no single summary.

A **design** is the answer: a named set of scenarios run as **one** pipeline,
unioned into one possibly-ragged task set. Overlapping cells are computed
**once**, and the result is a single combined summary with a `scenario` column.
This vignette shows how to grow a one-off run into a design --- a one-line
change --- and then refine it.

## Start with a single scenario

A typical one-off run defines a scenario and turns it into a `targets` pipeline
with `ssd_scenario_targets()`:

```{r}
#| label: scenario
data <- ssd_scenario_data(ssddata::ccme_boron)
coarse <- ssd_define_scenario(
  data,
  nsim = 10L,
  seed = 42L,
  nrow = c(5L, 10L, 20L),
  dists = ssd_distset(lnorm = "lnorm")
)
coarse
```

```{r}
#| label: scenario-targets
#| eval: false
# _targets.R
library(targets)
library(tarchetypes)
library(ssdsims)

ssd_scenario_targets(coarse, root = "results")
```

## Migrate: wrap it in a design

To grow this into a study, wrap the scenario with `ssd_design()` and switch the
factory to `ssd_design_targets()`. That is the entire change:

```{r}
#| label: design-one
design <- ssd_design(coarse)
design
```

```{r}
#| label: design-targets-one
#| eval: false
# _targets.R
design <- ssd_design(coarse)
ssd_design_targets(design, root = "results")
```

A design of **one** is valid and uniformly shaped --- the recommended starting
point for a study that may grow. The per-task results are **byte-identical** to
the standalone run: combining changes only *addressing* (target names and the
results tree), never a task's reproducible `(seed, primer)`.

::: callout-note
## Safe but recomputing

The design tree gains a `seed=<value>` level the standalone `layout=` tree
lacks (`results/seed=42/layout=.../...`), so the *first* design run recomputes
into the new tree. Nothing is re-baselined --- the numbers are identical --- it
is a one-time relocation. From then on, growing the design is incremental.
:::

## Refine: add a denser member

Now add the refinement. Suppose you want a finer `nrow` sweep --- but only the
extra sizes, not a whole new rectangular run. Define a second scenario covering
just the new region and add it to the design:

```{r}
#| label: design-refine
dense <- ssd_define_scenario(
  data,
  nsim = 10L,
  seed = 42L,
  nrow = c(12L, 14L, 16L, 18L),
  dists = ssd_distset(lnorm = "lnorm")
)
study <- ssd_design(coarse = coarse, dense = dense)
study
```

```{r}
#| label: design-targets-refine
#| eval: false
# _targets.R
ssd_design_targets(study, root = "results")
```

Both members share the same `seed`, dataset, and distributions, differing only
in their `nrow` coverage. So they **share** every `sample` shard (the draw does
not depend on `nrow`) and the `fit`/`hc` shard *cells* --- the union merges
their per-`nrow` tasks into the shared shards. Re-running `tar_make()` builds
only the genuinely new work; the cells `coarse` already computed stay cached.

The combined `results/summary.parquet` carries a `scenario` column (`"coarse"`
or `"dense"`), and each member's rows are filtered to exactly its own cells ---
ready to plot the coarse and refined sweeps together.

## Notes

- **Consistency contract.** `ssd_design()` requires that a name means the same
  thing across members: the same `dataset` name must bind identical data, the
  same `min_pmix` name the same function, the same `distset` name the same
  members, and `partition_by` must match. This is what makes sharing a cell
  across members sound. Inconsistent bindings abort at construction.
- **Varying the seed.** Members *may* use different `seed`s (e.g. repeating the
  whole exploration under several master seeds); they land under separate
  `seed=` trees and share nothing. Members sharing a `seed` share their
  coincident cells (common random numbers).
- **Keep `nrow_max` uniform.** `nrow_max` is a sample draw-size guard, not a
  comparison axis; differing or changing it across members is undefined
  behaviour for shard sharing.

See ["Running a Sharded Pipeline"](sharded-pipeline.html) for the shard/results
layout a design builds on.
