---
title: "Running a Sharded Pipeline"
vignette: >
  %\VignetteIndexEntry{Running a Sharded Pipeline}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
# The live chunks materialise Parquet shards, so they need the fitting deps and
# duckplyr. Skip evaluation gracefully on a minimal runner rather than failing.
evaluate <- requireNamespace("ssdsims", quietly = TRUE) &&
  requireNamespace("ssddata", quietly = TRUE) &&
  requireNamespace("ssdtools", quietly = TRUE) &&
  requireNamespace("duckplyr", quietly = TRUE)
knitr::opts_chunk$set(eval = evaluate)
```

```{r}
#| label: library
library(ssdsims)
```

The `vignette("defining-a-scenario")` article ends
with the in-memory baseline runner, which threads each step's results forward in
memory. This vignette covers the next layer: materialising each step as
**Hive-partitioned Parquet shards** and linking the steps by reading parent
shards back from disk --- the storage hand-off the cluster
[targets](https://docs.ropensci.org/targets/) pipeline is built on.

The key idea is **two drivers over one execution core**. The per-task work
(draw, fit, hazard concentration) and its reproducible per-task RNG are shared;
only the *I/O and granularity* differ:

- `ssd_run_scenario_baseline()` --- single core, results threaded in memory.
- `ssd_run_scenario_shards()` --- single core, results written as Parquet shards
  and read back by name (**this vignette**).
- a `targets` pipeline --- the same shards, one named target each, runnable in
  parallel / on a cluster (**shown but not run below**).

Because each task installs its own `(seed, primer)`, all three produce
**byte-identical** per-task results: the storage layout is a free re-layout.

## `partition_by` and `bundle`: how tasks group into shards

A **shard** is one Parquet file: one cell of the `partition_by` *path* axes for
a step. The remaining axes --- the `bundle` (inner) axes --- become ordinary
columns *within* a shard, so several tasks share one file. The two are exact
complements (`path ⊎ inner = task_axes(step)`), and you can specify either,
per step, with at most one of `partition_by`/`bundle` naming a given step.

```{r}
#| label: scenario
data <- ssd_scenario_data(ssddata::ccme_boron)
scenario <- ssd_define_scenario(
  data,
  nsim = 2L,
  seed = 42L,
  nrow = c(5L, 10L),
  rescale = c(FALSE, TRUE),
  dists = ssd_distset(set = c("lnorm", "gamma")))
scenario
```

The `print()` shows, per step, the `partition_by` (across-shards) path axes and
the `bundle` (within-shard) inner axes. The default coarsens downstream:
`fit` shards on `(dataset, sim, nrow, rescale)`, while `hc` shards only on
`(dataset, sim)`. Steps partition **independently** --- there is no cross-step
constraint --- so an `hc` shard reads *several* `fit` shards and a `fit` shard
can read *several* `sample` shards. That **m:n** relationship is resolved when
the parent shards are read back, not by restricting `partition_by`.

The stored `partition_by` is the per-step path list (the `bundle`/inner axes
shown in `print()` are its complement):

```{r}
#| label: axes
scenario$partition_by$fit
```

### `distset`: bundled by default, promotable to the path

When a scenario declares several distribution sets (pools carved from one union
fit, see the *defining-a-scenario* vignette), `distset` --- the set *name* --- is
an `hc` axis. By default it is an **inner** (bundled) axis: the `hc` path stays
`c("dataset", "sim")`, so **one** hc shard holds every pool for a
`(dataset, sim)` cell, the runner decodes the union fit *once*, and subsets it
N ways in memory. This keeps shard counts sane (35 datasets × 1000 sims × 7
pools would otherwise be 245k tiny Parquets):

```{r}
#| label: distset-bundled
multi <- ssd_define_scenario(
  data,
  nsim = 2L,
  seed = 42L,
  dists = ssd_distset(
    BCANZ = ssdtools::ssd_dists_bcanz(),
    Iwasaki = c("burrIII3", "gamma", "llogis", "lnorm", "weibull"),
    lnorm = "lnorm"
  )
)
# distset is an inner (bundled) hc axis by default
multi$partition_by$hc
nrow(ssd_scenario_hc_shards(multi)) # one hc shard per (dataset, sim) cell
```

Promote `distset` to a path axis (one shard per pool per cell) when you want
per-pool shards --- e.g. to cache and re-upload pools independently, or because a
later-added pool should mint only its own new hc shards while every existing
shard stays cached (the fit layer carries no `distset`, so its shards are never
touched):

```{r}
#| label: distset-path
per_pool <- ssd_define_scenario(
  data,
  nsim = 2L,
  seed = 42L,
  dists = ssd_distset(
    BCANZ = ssdtools::ssd_dists_bcanz(),
    lnorm = "lnorm"
  ),
  partition_by = list(hc = c("dataset", "sim", "distset"))
)
nrow(ssd_scenario_hc_shards(per_pool)) # one hc shard per (dataset, sim, distset)
```

## Run it single core, over shards

`ssd_run_scenario_shards()` runs the three steps and writes one Parquet per
shard under a Hive-partitioned tree, with **no** `targets` dependency. It
returns the shard paths per step.

```{r}
#| label: run
run <- ssd_run_scenario_shards(scenario)
run
```

The written tree is Hive-partitioned (`<step>/<axis=value>/.../part.parquet`),
so the path itself identifies the shard:

```{r}
#| label: layout
fit_files <- list.files(
  file.path(run$dir, "fit"),
  pattern = "part.parquet",
  recursive = TRUE
)
fit_files
```

One file per `partition_by` path cell --- here `1 (dataset) x 2 (sim) x 2 (nrow)
x 2 (rescale) = 8` fit shards --- not one per task. Each file carries the step's
inner axes and the per-task results as columns.

### Results match the in-memory runner

`partition_by` is a free re-layout, so the sharded run's per-task results equal
the in-memory baseline's. The `hc` estimates, read back from the shards, line up
with `ssd_run_scenario_baseline()` joined on the task identity:

```{r}
#| label: oracle
base <- ssd_run_scenario_baseline(scenario)
base_est <- sort(do.call(rbind, base$hc$hc)$est)

summary_path <- ssd_summarise(
  file.path(run$dir, "sample"),
  file.path(run$dir, "fit"),
  file.path(run$dir, "hc"),
  file.path(run$dir, "summary.parquet")
)
summary_tbl <- dplyr::collect(duckplyr::read_parquet_duckdb(summary_path))
shard_est <- sort(summary_tbl$est)

all.equal(base_est, shard_est)
```

`ssd_summarise()` fans the `hc` layer in across shards (via duckplyr) into a
single `summary.parquet` --- the analysis-ready table --- without recomputing
anything. It projects the `dists`/`samples` list-columns out at the DuckDB level,
so the (potentially large) retained bootstrap draws are never pulled into R. Pass
`path_with_samples =` to **also** write a full summary that keeps those columns; that
write stays lazy in DuckDB, so the draws never materialise in R there either. In
the `targets` pipeline the full `summary-samples.parquet` is written
automatically when the scenario set `samples = TRUE` (the case where the retained
draws carry information the compact summary cannot).

## Scaling up: the `targets` pipeline

The sharded runner is single core. To run shards **in parallel** (or on a
cluster), the same building blocks plug into a static-branching `targets`
pipeline: `tarchetypes::tar_map()` mints one named target per shard, so
`targets` caches, invalidates, and reruns each shard independently. Three ready
example projects ship with the package --- a minimal `small` one, a fuller
`large` one (adapted from `scripts/example.R`, sweeping `nrow`, `proportion`,
and the estimation / CI methods), and a `cluster` one for SLURM. The `large`
project parallelises its shards across local workers with a mirai-backed `crew`
controller (`tar_option_set(controller = crew::crew_controller_local())`); the
`cluster` project is the same pipeline under a
`crew.cluster::crew_controller_slurm()` controller, with a connectivity +
worker-prerequisite preflight — see
`vignette("cluster-pipeline")`, a "zero to a running
cluster job" guide that maps your site's own SLURM instructions onto the
controller:

```{r}
#| label: template-path
list.files(system.file("targets-templates", "small", package = "ssdsims"))
```

The `small` and `large` directories each hold `scenario.R` (the study, shared by
both drivers), `_targets.R` (the pipeline), `run.R` (the **targets** driver), and
`run-serial.R` (the **single-core** driver); the `cluster` directory instead
inlines the scenario in `_targets.R` and adds a `controller.R` (the SLURM
controller), a `preflight.R` (connectivity check), a `run.R` driver, and a
`README.md` guide. The whole `_targets.R` is just
*build a scenario and call the `ssd_scenario_targets()` factory* —

```{r}
#| label: factory
#| eval: false
library(targets)
library(tarchetypes)
library(ssdsims)
source("scenario.R") # builds `scenario`
ssd_scenario_targets(scenario) # the entire target list
```

— so editing the study means editing `scenario.R` only. Copy a directory's files
to your project root, edit `scenario.R`, then run either driver (the targets
path needs the `targets`/`tarchetypes` Suggests):

```{r}
#| label: tar-make
#| eval: false
# install.packages(c("targets", "tarchetypes"))
dir <- system.file("targets-templates", "small", package = "ssdsims")
file.copy(list.files(dir, full.names = TRUE), ".")

source("run.R") # targets: tar_make() -> one target per shard -> results/<step>/...
# or, from a shell:  Rscript run.R

source("run-serial.R") # single core via ssd_run_scenario_shards() -> results-serial/
# ...and, if results/ exists, it checks (and prints) whether the two drivers' estimates match
```

`ssd_scenario_targets()` mints one `format = "file"` target per shard (targets
passes the shard *path*, not its value, between steps) with `error = "null"` (a
failing whole shard records its error and goes `NULL` without aborting the run
--- the rest still build and the summary unions the survivors), and wires step
ordering with `tar_combine()` barriers (a directory read carries no automatic
dependency edge).

Because a shard's Hive-path depth depends on `partition_by`/`bundle`, replaying
with a *changed* split must not leave stale-granularity shards beside the new
ones (a `**` glob would union both). The two drivers handle this differently:
`run.R` (targets) writes each layout under its own root,
`scenario_results_dir(scenario)` = `results/seed=<value>/layout=<hash>/`, so a
split change is a fresh tree; `run-serial.R` (single core) instead **owns** its
`dir` and clears each step subtree on every run.

Because both drivers `source("scenario.R")` --- the same study --- and
`partition_by` is a free re-layout, `run-serial.R` finishes by reading back its
own `summary.parquet` and the targets `scenario_results_dir(scenario)` summary
and checking (with `all.equal()`) that the per-task estimates match. Whatever the
driver --- baseline, single-core shards, or `targets` --- the results are the
same; choose the one that fits the run.

## Combining scenarios into a design

A single scenario is one **regular** grid. To explore finer detail over a
*subset* of the axes --- a coarse grid everywhere plus a dense refinement in one
region --- without computing the full cross-product, union several scenarios into
a **design** and run them as one pipeline with `ssd_design_targets()`. Members
sharing a `seed` share their coincident shards (computed once), and the design
fans in one combined `summary.parquet` with a `scenario` identity column. Growing
a one-off run into a design is a one-line switch; see
`vignette("scenario-to-design")`.

### Comparing settings in one design

A design also serves the *secondary* use of comparing **settings** at the same
cells. Members of a seed group may differ in the four non-axis hc readouts
(`proportion`, `est_method`, `ci`, `samples`) and in their fit `dists` union ---
only the layout-shaping `nrow_max` and `partition_by` must agree. Differing
readouts are reconciled **per shared hc cell**: `proportion`/`est_method` are
unioned and `ci`/`samples` reduced with `any()`, so the cell computes the maximal
readout set in one shard and each member's summary filters its own slice.

This is what makes a cheap-versus-expensive comparison economical. Put a
broad `ci = FALSE` member beside a narrow `ci = TRUE` one --- say a point-estimate
sweep over many sims and a confidence-interval check over a handful:

```r
broad  <- ssd_define_scenario(data, nsim = 1000L, seed = 42L, ci = FALSE)
narrow <- ssd_define_scenario(data, nsim = 10L,  seed = 42L, ci = TRUE, nboot = 1000L)
ssd_design_targets(ssd_design(broad = broad, narrow = narrow))
```

The expensive bootstrap runs **only** on the 10 overlapping sims (where `narrow`
has tasks); the other 990 stay `ci = FALSE`. The `ci = FALSE` member reads its
(analytical, ci-invariant) point `est` for the overlapping sims straight from the
`ci = TRUE` shards rather than recomputing them. Members differing only in their
`distset` coverage likewise share every `sample`/`fit` shard (one design-wide
union fit, each member subsetting via its `distset` axis) and differ only in their
`distset` hc cells. In every case a member's per-task results are byte-identical
to its standalone run.

## See also

- `vignette("defining-a-scenario")` --- the scenario
  object and the baseline runner.
- `vignette("scenario-to-design")` --- union
  scenarios into one ragged pipeline that shares overlapping shards.
- `vignette("cluster-pipeline")` --- the same pipeline
  on a cluster, and `vignette("cloud-upload")` --- shipping
  the shards to an object store so they are readable off the cluster.
- `ssd_run_scenario_shards()`,
  `ssd_scenario_fit_shards()`,
  `ssd_run_fit_step()`,
  `ssd_summarise()`.
